The Sloan Digital Sky Survey Spectroscopic Reduction Pipeline

### 2006-03-31 doc/paper/paper2d.tex
updated observations section

@@

We utilize an optimal extraction procedure to transform a 2-dimensional image of
the fiber spectrograms to a set of 320 1-dimensional spectra (per spectrograph).
We use the same procedure on all processed frames, 
and will describe the details in processing the flat-field frames, the
arc lamp frames, and finally the science frames.  
The extraction is performed row-by-row, so each frame requires an 
outside loop over 2048 individual rows.
The counts in each row of 2048 columns are modeled with a linear 
combination of 320 profiles plus a low-order
polynomial background.  In addition, deviations to the trace centers and
the profile widths are calculated as linear basis modes (reflecting the first
and second spatial derivatives, respectively).  The compiled routine
we implement to fit the profiles in a single row (extract\_row.c) can
be invoked with any combination of these parameters.

@@

The first step in object extraction is a simple tophat aperture (boxcar) 
extraction centered at the fiber positions determined from the flat-field 
and with a full spatial width of 6 CCD pixels.  The median counts per fiber 
are passed directly to a test for overly bright fibers. Fiber spectra are 
identified as bright (a.k.a. "WHOPPING") if the median counts per pixel in the 
boxcar extraction are greater than 10000 counts above the median in the 
frame.  In addition, only the brightest fiber in a spatially contiguous set of 
bright fibers is labelled "WHOPPING".

@@

@@

>>>>

\section{Typical Spectroscopic Observations}
\label{sec_observ}

At the beginning of a typical survey night, up to 9 plates have been mounted
in cartridges, plugged with 640 science fibers and 11 guide fibers. 
The position of the 640 fibers along each of the two fiber slitheads
are mapped by the 8th wonder of the world,
the SDSS plugplate fiber mapper (Schlegel and Finkbeiner).
All usable spectroscopic data obtained for a single plugging, 
and therefore a single fiber mapping, will eventually be combined into 
a final data product for that plugging.  In this scheme, an individual 
output spectrum is uniquely specified by the plate number, the first night
of observations (in MJD), and the science fiber number.

A plate is mounted in the telescope through a efficient procedure described
in Gunn et al. 2006.  The switch from the imaging camera to a spectroscopic
cartridge takes an average of xx minutes, while a switch from one cartridge
to another takes merely x minutes.  Once the cartridge is mounted, the telescope is slewed to the coordinates of the plate center.  
Before acquisition of the guide stars, the 8 flat-field petals are closed and
a sequence of one 10-second flat-field exposure and one 2-second 
Neon-Mercury-Cadmium arc exposure are obtained.  While the arc-lamp frame is
being readout, the flat-field petals are opened, and guide stars are acquired
through the 11 coherent guide bundles (9 are this big and 2 are even bigger??).
With at least 4 guide stars acquired, closed loop guiding is started, and
usually converges to stable guiding within 2 minutes??
The plate scale of the SDSS telescope can be adjusted, and the guiding software
suggest a scale change if one is required.  Once close loop 
guiding has settled, the first science exposure is started.  The typical 
exposure time is 900 seconds, but exposure times from 600 to 1200 second are
not unexpected based on conditions and time constraints.  A second exposure
is started immediately after the first science exposure is complete and readout.A fast linux workstation onsite copies the data from the data acquisition host,
and automatically performs a quick spectral extraction with a subset of the 
steps explained here.  The results are posted to sos.apo.nmsu.edu, and
the SDSS observers can check the S/N obtained in each of the 4 spectroscopic
cameras in real-time.  A plate is considered to have met SDSS survey standards
if all 4 cameras have a cumulative (S/N)^2 greater than 15 at 
fidicual magnitudes of ?? and ??.

@@

We utilize an optimal extraction procedure to transform a 2-dimensional image of
the fiber spectrograms to a set of 320 1-dimensional spectra (per spectrograph).
We use the same procedure on all processed frames, 
and will describe the details in processing the flat-field frames, the
arc lamp frames, and finally the science frames.  
The extraction is performed row-by-row, so each frame requires an 
outside loop over 2048 individual rows.
The counts in each row of 2048 columns are modeled with a linear 
combination of 320 profiles plus a low-order
polynomial background.  In addition, deviations to the trace centers and
the profile widths are calculated as linear basis modes (reflecting the first
and second spatial derivatives, respectively).  The compiled routine
we implement to fit the profiles in a single row (extract\_row.c) can
be invoked with any combination of these parameters.  The parameters are
sorted in order of fiber number with background parameters included at the
end of the list.  This facilitates and banded matrix inversion, and 
significantly increases the speed of Cholesky decomposition step in each
iteration of the extraction.

@@

The first step in object extraction is a simple tophat aperture (boxcar) 
extraction centered at the fiber positions determined from the flat-field 
and with a full spatial width of 6 CCD pixels.  The median counts per fiber 
are passed directly to a test for overly bright fibers. Fiber spectra are 
identified as bright (a.k.a. "WHOPPING") if the median counts per pixel in the 
boxcar extraction are greater than 10000 counts above the median in the 
frame.  In addition, only the brightest fiber in a spatially contiguous set of 
bright fibers is labelled "WHOPPING".  Each whopping fiber adds an additional
background parameter to the fit per row, which is an exponential profile
with a spatial scale of 25 pixels.  Such a profile correlates with many
neighboring fibers, so the parameter is treated as a background to supplement
the smooth polynomial background.  The flux contained in the exponential 
profile is not included in the extracted spectra.

@@

%------------------------------------------------------------------------------
% ACKNOWLEDGMENTS
%------------------------------------------------------------------------------
\bigskip
We are very grateful to Michael Strauss for
useful and helpful comments on the manuscript.
Funding for the creation and distribution of the SDSS Archive has been
provided by the Alfred P. Sloan Foundation, the Participating
Institutions, the National Aeronautics and Space Administration, the
National Science Foundation, the U.S. Department of Energy, the
Japanese Monbukagakusho, and the Max Planck Society. The SDSS Web site
is http://www.sdss.org/. The SDSS is managed by the Astrophysical
Research Consortium (ARC) for the Participating Institutions. The
Participating Institutions are The University of Chicago, Fermilab,
the Institute for Advanced Study, the Japan Participation Group,
The Johns Hopkins University, the Korean Scientist Group, Los Alamos
National Laboratory, the Max-Planck-Institute for Astronomy (MPIA),
the Max-Planck-Institute for Astrophysics (MPA), New Mexico State
University, University of Ohio, University of Pittsburgh,
University of Portsmouth,
Princeton University, the United States Naval Observatory, and
the University of Washington.

@@

\bibitem[Wild \& Hewett(2005)]{2005MNRAS.358.1083W} Wild, V., \& Hewett, 
P.~C.\ 2005, \mnras, 358, 1083
